# Reproducible Research: Peer Assessment 1 by Mohan Suswaram

## Loading and preprocessing the data

```{r}
    ## unzip the data set
    unzip("activity.zip")
    
    # read the data
    data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

```{r}
    # find the total number of steps taken each day
    steps.per.day <- aggregate(steps~date, data=data, FUN=sum)
```
```{r fig.width=6, fig.height=6}
    # make a histogram of the total number of steps taken each day
    hist(steps.per.day$steps, breaks=nrow(steps.per.day), main="Total Number of Steps Per Day",
         xlab="Steps Per Day", col="red")
```
```{r}
    # Calculate the mean and median total number of steps taken per day
    mean.steps.per.day <- mean(steps.per.day$steps)
    median.steps.per.day <- median(steps.per.day$steps)
```
The mean total number of steps taken per day is `r format(mean.steps.per.day, scientific=F)`  
The median total number of steps taken per day is `r format(median.steps.per.day, scientific=F)`

## What is the average daily activity pattern?

```{r}
    # find the average number of steps taken per 5 minute interval
    average.per.interval <- aggregate(steps~interval, data=data, FUN=mean)
```
```{r fig.width=6, fig.height=6}
    # make a time series plot of the 5-minute interval and the 
    # average number of steps taken, averaged across all days
    plot(average.per.interval, type="l", main="Average Daily Activity Pattern",
         xlab="5-minute Intervals Over Day", ylab="Average Steps Taken Over All Days")
```
```{r}
    # find the 5-minute interval, on average across all the days in the dataset, 
    # that contains the maximum number of steps
    max.interval <- average.per.interval[which.max(average.per.interval$steps),"interval"]
```
The 5-minute interval, on average across all days, that contains the maximum number of steps is `r max.interval`

## Imputing missing values
```{r}
    ## calculate the total number of missing values in the dataset 
    total.missing <- sum(is.na(data))
```
The total number of missing values in the dataset is `r total.missing`.  
  
The missing values will be imputed by replacing step NAs with the mean of the 5-minute interval averaged across all days.

```{r}

    # create a vector of steps with NAs replaced by imputed value (mean of 5-minute interval)
    imputed.steps <- numeric()
    for(i in 1:nrow(data)) {
        obs <- data[i,]
        if (is.na(obs$steps)) {
            steps <- subset(average.per.interval,interval==obs$interval)$steps
        } else {
            steps <- obs$steps
        }
        imputed.steps <- c(imputed.steps,steps)
    }
    
     ## create a new dataset that is equal to the original dataset but with the missing data filled in.
    imputed.data <- data
    imputed.data$steps <- imputed.steps
    
    # find the total number of steps taken each day
    imputed.steps.per.day <- aggregate(steps~date, data=imputed.data, FUN=sum)
```
```{r fig.width=6, fig.height=6}
    # make a histogram of the total number of steps taken each day
    hist(imputed.steps.per.day$steps, breaks=nrow(imputed.steps.per.day), 
         main="Total Number of Steps Per Day With Imputed Values",
         xlab="Steps Per Day", col="blue")
```
```{r}
    # Calculate the mean and median total number of steps taken per day
    imputed.mean.steps.per.day <- mean(imputed.steps.per.day$steps)
    imputed.median.steps.per.day <- median(imputed.steps.per.day$steps)
```
The mean total number of steps taken per day is `r format(imputed.mean.steps.per.day, scientific=F)`  
The median total number of steps taken per day is `r format(imputed.median.steps.per.day, scientific=F)`

After imputing the missing step values, the mean total number of steps per day remained unchanged while the median total number of steps per day changed from `r format(median.steps.per.day, scientific=F)` to `r format(imputed.median.steps.per.day, scientific=F)`. The impact of the imputation was
a slight increase in the median total number of steps per day.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
    ## change date column from factor to Date
    
    imputed.data$date <- as.Date(imputed.data$date)

    ## create a new factor variable in the dataset with two levels 
    ## “weekday” and “weekend” indicating whether date is a weekday or weekend day.

    weekend.days <- c("Saturday","Sunday")
    imputed.data$daytype <- as.factor(sapply(imputed.data$date, function(x) ifelse(weekdays(x) %in% weekend.days,"weekend","weekday")))
```
```{r fig.width=6, fig.height=6}
    ## Make a panel plot containing a time series plot (i.e. type = "l") 
    ## of the 5-minute interval (x-axis) and the average number of steps taken, 
    ## averaged across all weekday days or weekend days (y-axis).
    require(plyr)
    average.steps <- ddply(imputed.data, .(interval, daytype), summarize, steps = mean(steps))

    require(lattice)
    xyplot(steps ~ interval | daytype, data = average.steps, layout = c(1, 2), type = "l", 
     xlab="5-minute Intervals Over Day", ylab="Number of Steps",
     main="Activity Patterns on Weekends and Weekdays")
```